# Appendix: Bookdown and LaTeX Notes{.unnumbered #latex}

Ray's Formatting Notes: version 2021-09-22
https://gist.github.com/RaymondBalise/ee4b7da0a70087317dc52bf479a4e2b6

```{=html}
<style>
  .col2 {
    columns: 2 150px;         /* number of columns and width in pixels*/
    -webkit-columns: 2 150px; /* chrome, safari */
    -moz-columns: 2 150px;    /* firefox */
  }
  .col3 {
    columns: 3 100px;
    -webkit-columns: 3 100px;
    -moz-columns: 3 100px;
  }
</style>
```

## Markdown highlighting {.unnumbered}

| Formatting | Code         |
|:-----------|:-------------|
| **bold**   | \*\*bold\*\* |
| __bold__   | \_\_bold\_\_ |
| *italic*   | \*italic\*   |
| _italic_   | \_italic\_   |

## Text coloring {.unnumbered}

To add color you can use CSS or R code like this:

```{r, include=TRUE}
color_text <- function(x, color){
  if(knitr::is_latex_output())
    paste("\\textcolor{",color,"}{",x,"}",sep="")
  else if(knitr::is_html_output())
    paste("<font color='",color,"'>",x,"</font>",sep="")
  else
    x
}

red <- function(x){
  if(knitr::is_latex_output())
    paste("\\textcolor{",'red',"}{",x,"}",sep="")
  else if(knitr::is_html_output())
    paste("<font color='red'>",x,"</font>",sep="")
  else
    x
}
```

The strng `r red("This is colored with the red function")` comes from:

```{r eval=FALSE}
`r red("This is colored with the red function")`
```

The string `r color_text("This is colored with the color_text function set to mauve", "#E0B0FF")` comes from:

```{r eval=FALSE}
`r color_text("This is colored with the color_text function set to mauve", "#E0B0FF")`
```

## Section references {.unnumbered #x99-4}

Section [99.4](#x99-4) comes from `Section [99.4](#x99-4)`

## Footnotes {.unnumbered}

^[A footnote] comes from `^[A footnote]`

## Formatting Text {.unnumbered}

| Appearance                                               | Code                                                   |
|:---------------------------------------------------------|:-------------------------------------------------------|
| `r xfun::n2w(3, cap = TRUE)`                             | `xfun::n2w(3, cap = TRUE)`                             |
| `r scales::percent(0.1447, accuracy= .1)`                | `scales::percent(0.1447, accuracy= .1)`                |
| `r scales::pvalue(0.1447, accuracy= .001, add_p = TRUE)` | `scales::pvalue(0.1447, accuracy= .001, add_p =TRUE)`  |

## Figures {.unnumbered}

Name a figure chunk with a name like figure99-1 and include fig.cap="something" then reference it like this: `\@ref(fig:figure99-1)`

## Displaying Formula {.unnumbered}

### Formatting {.unnumbered}

To tweak the appearance of words use these formats:

| Formatting        | Code                    | Looks like              |
|:------------------|:------------------------|-------------------------|
| plain text        | \\text{text Pr}         | $\text{text Pr}$        |
| bold Greek symbol | \\boldsymbol{\\epsilon} | $\boldsymbol{\epsilon}$ |
| typewriter        | \\tt{blah}              | $\tt{blah}$             |
| slide font        | \\sf{blah}              | $\sf{blah}$             |
| bold              | \\mathbf{x}             | $\mathbf{x}$            |
| plain             | \\mathrm{text Pr}       | $\mathrm{text Pr}$      |
| cursive           | \\mathcal{S}            | $\mathcal{S}$           |
| Blackboard bold   | \\mathbb{R}             | $\mathbb{R}$            |

### Symbols {.unnumbered}

| Symbols                    | Code                       |
|:---------------------------|:---------------------------|
| $\stackrel{\text{def}}{=}$ | \\stackrel{\\text{def}}{=} |

### Notation {.unnumbered}

Based on: <https://www.calvin.edu/~rpruim/courses/s341/S17/from-class/MathinRmd.html>

| Math                                                          | Code                                                               |
|:--------------------------------------------------------------|:-------------------------------------------------------------------|
| $x = y$                                                       | `$x = y$`                                                          |
| $x \approx y$                                                 | `$x \approx y$`                                                    |
| $x < y$                                                       | `$x < y$`                                                          |
| $x > y$                                                       | `$x > y$`                                                          |
| $x \le y$                                                     | `$x \le y$`                                                        |
| $x \ge y$                                                     | `$x \ge y$`                                                        |
| $x \ge y$                                                     | `$x \ge y$`                                                        |
| $x \times y$                                                  | `$x \times y$`                                                     |
| $x^{n}$                                                       | `$x^{n}$`                                                          |
| $x_{n}$                                                       | `$x_{n}$`                                                          |
| $\overline{x}$                                                | `$\overline{x}$`                                                   |
| $\hat{x}$                                                     | `$\hat{x}$`                                                        |
| $\widehat{SE}$                                                | `$\widehat{SE}$`                                                   |
| $\tilde{x}$                                                   | `$\tilde{x}$`                                                      |
| $\frac{a}{b}$                                                 | `$\frac{a}{b}$`                                                    |
| $\displaystyle \frac{a}{b}$                                   | `$\displaystyle \frac{a}{b}$`                                      |
| $\binom{n}{k}$                                                | `$\binom{n}{k}$`                                                   |
| $x_{1} + x_{2} + \cdots + x_{n}$                              | `$x_{1} + x_{2} + \cdots + x_{n}$`                                 |
| $x_{1}, x_{2}, \dots, x_{n}$                                  | `$x_{1}, x_{2}, \dots, x_{n}$`                                     |
| $\mathbf{x} = \langle x_{1}, x_{2}, \dots, x_{n}\rangle$      | `$\mathbf{x} = \langle x_{1}, x_{2}, \dots, x_{n}\rangle$`         |
| $x \in A$                                                     | `$x \in A$`                                                        |
| $|A|$                                                         | `$|A|$`                                                            |
| $x \in A$                                                     | `$x \in A$`                                                        |
| $x \subset B$                                                 | `$x \subset B$`                                                    |
| $x \subseteq B$                                               | `$x \subseteq B$`                                                  |
| $A \cup B$                                                    | `$A \cup B$`                                                       |
| $A \cap B$                                                    | `$A \cap B$`                                                       |
| $X \sim {\sf Binom}(n, \pi)$                                  | `X \sim {\sf Binom}(n, \pi)$`                                      |
| $\mathrm{P}(X \le x) = {\tt pbinom}(x, n, \pi)$               | `$\mathrm{P}(X \le x) = {\tt pbinom}(x, n, \pi)$`                  |
| $P(A \mid B)$                                                 | `$P(A \mid B)$`                                                    |
| $\mathrm{P}(A \mid B)$                                        | `$\mathrm{P}(A \mid B)$`                                           |
| $\{1, 2, 3\}$                                                 | `$\{1, 2, 3\}$`                                                    |
| $\sin(x)$                                                     | `$\sin(x)$`                                                        |
| $\log(x)$                                                     | `$\log(x)$`                                                        |
| $\int_{a}^{b}$                                                | `$\int_{a}^{b}$`                                                   |
| $\left(\int_{a}^{b} f(x) \; dx\right)$                        | `$\left(\int_{a}^{b} f(x) \; dx\right)$`                           |
| $\left[\int_{-\infty}^{\infty} f(x) \; dx\right]$             | `$\left[\int_{\-infty}^{\infty} f(x) \; dx\right]$`                |
| $\left. F(x) \right|_{a}^{b}$                                 | `$\left. F(x) \right|_{a}^{b}$`                                    |
| $\sum_{x = a}^{b} f(x)$                                       | `$\sum_{x = a}^{b} f(x)$`                                          |
| $\prod_{x = a}^{b} f(x)$                                      | `$\prod_{x = a}^{b} f(x)$`                                         |
| $\lim_{x \to \infty} f(x)$                                    | `$\lim_{x \to \infty} f(x)$`                                       |
| $\displaystyle \lim_{x \to \infty} f(x)$                      | `$\displaystyle \lim_{x \to \infty} f(x)$` \`                      |
| $RMSE = \sqrt{\frac{1}{n}\sum_{i=1}^{n} (Y_n - \hat{Y}_i)^2}$ | `$RMSE = \sqrt{\frac{1}{n}\sum_{i=1}^{n} (Y_n - \hat{Y}_i)^2}$` \` |

## Equations {.unnumbered}

These are formulas that appear with an equation number.

### Basic Equation {.unnumbered}

The names of equations can not include . or \_ but it can include -

    \begin{equation} 
      1 + 1 = 2 
      (\#eq:eq99-1)
    \end{equation} 

Which appears as:

\begin{equation} 
  1 + 1 = 2
  (\#eq:eq99-1)
\end{equation} 

The reference to the equation is \@ref(eq:eq99-1) which comes from this code `\@ref(eq:eq99-1)`

### Case-When Equation (Large Curly Brace) {.unnumbered}

Based on: <https://tex.stackexchange.com/questions/9065/large-braces-for-specifying-values-of-variables-by-condition> 

Case when formula: 

```{=tex}
\begin{equation} 
y =
\begin{cases}
  0, & \text{if}\ a=1 \\
  1, & \text{otherwise}
\end{cases}
  (\#eq:eq99-2)
\end{equation} 
```

Which comes from this code:

```{}
    \begin{equation} 
      y =
      \begin{cases}
        0, & \text{if}\ a=1 \\
        1, & \text{otherwise}
      \end{cases}
      (\#eq:eq99-2)
    \begin{equation} 
```

The reference to equation is \@ref(eq:eq99-2) which comes from this code `\@ref(eq:eq99-2)`

### Alligned with Underbars {.unnumbered}

\begin{equation}
\begin{aligned}
\mathrm{E}(Y-\hat{Y})^2 & = \mathrm{E}[f(X) + \epsilon -\hat{f}(X)]^2 \\
                        & = \underbrace{[f(X) -\hat{f}(X)]^2}_{\mathrm{Reducible}} + \underbrace{\mathrm{var}(\epsilon)}_{\mathrm{Irreducible}} \\
\end{aligned}
(\#eq:eq99-3)
\end{equation}

Comes from this code:

```{}
\begin{equation}
\begin{aligned}
\mathrm{E}(Y-\hat{Y})^2 & = \mathrm{E}[f(X) + \epsilon -\hat{f}(X)]^2 \\
                        & = \underbrace{[f(X) -\hat{f}(X)]^2}_{\mathrm{Reducible}} + \underbrace{\mathrm{var}(\epsilon)}_{\mathrm{Irreducible}} \\
\end{aligned}
(\#eq:eq99-3)
\end{equation}
```

## Greek letters {.unnumbered}

Based on: <https://www.calvin.edu/~rpruim/courses/s341/S17/from-class/MathinRmd.html>

::: {.col2}
| letters                   | code                        |
|:--------------------------|:----------------------------|
| $\alpha A$                | `$\alpha A$`                |
| $\beta B$                 | `$\beta B$`                 |
| $\gamma \Gamma$           | `$\gamma \Gamma$`           |
| $\delta \Delta$           | `$\delta \Delta$`           |
| $\epsilon \varepsilon E$  | `$\epsilon \varepsilon E$`  |
| $\zeta Z \sigma$          | `$\zeta Z \sigma`           |
| $\eta H$                  | `$\eta H$`                  |
| $\theta \vartheta \Theta$ | `$\theta \vartheta \Theta$` |
| $\iota I$                 | `$\iota I$`                 |
| $\kappa K$                | `$\kappa K$`                |
| $\lambda \Lambda$         | `$\lambda \Lambda$`         |
| $\mu M$                   | `$\mu M$`                   |
| $\nu N$                   | `$\nu N$`                   |
| $\xi\Xi$                  | `$\xi\Xi$`                  |
| $o O$                     | `$o O$ (omicron)`           |
| $\pi \Pi$                 | `$\pi \Pi$`                 |
| $\rho\varrho P$           | `$\rho\varrho P$`           |
| $\sigma \Sigma$           | `\sigma \Sigma$`            |
| $\tau T$                  | `$\tau T$`                  |
| $\upsilon \Upsilon$       | `$\upsilon \Upsilon$`       |
| $\phi \varphi \Phi$       | `$\phi \varphi \Phi$`       |
| $\chi X$                  | `$\chi X$`                  |
| $\psi \Psi$               | `$\psi \Psi$`               |
| $\omega \Omega$           | `$\omega \Omega$`           |
:::
